Scalar Feedback based Joint Time-Frequency Precoder interpolation for MIMO-OFDM Systems

### 2020-01-09 paper/main.tex
Final changes

An underlying challenge in framing the predictive quantization and interpolation algorithms has been the high dimensional unitary structure of the SVD precoders. A large body of works utilizes the manifold geometry endowed by the unitary structure to frame such algorithms directly on the non-linear manifold ~\cite{mondal2007quantization,schwarz2013adaptive,5946308,6891198,Gupt1905:Predictive,pitaval2013codebooks}. On the other hand, there are numerous works which utilize scalar parametrization of the unitary precoders, given by Givens rotations and Householder transformations \cite{4114278,4556174,lou2013comparison}. While the manifold based methods offer a clean solution and have been shown to be very effective in reducing the feedback overhead, they are analytically terse and
involve operations over higher dimensional manifolds. The scalar parametrization algorithms enable usage of the traditional algorithms to predict, interpolate and quantize over vector spaces, and hence are analytically simpler.
Such methods have also been implemented in 802.11ac standards
\cite{lou2013comparison}. We are not aware of performance comparisons
between these two bodies of works, thereby making it difficult to
quantify from past work whether performance gains/losses can accrue by operating over scalar parametrization instead of manifolds. In this paper, we propose a joint time-frequency interpolation approach that builds upon the existing parametric quantization techniques. One of the key results of our work is that the proposed scalar parametrization method actually outperforms an equivalent manifold based technique \cite{Gupt1905:Predictive}, at the same time being analytically more tractable.

@@

\section{Scalar joint time-frequency scheme}
\subsection{Scalar Parametrization}
\label{givens}
The degrees of freedom in the $\tilde{\bV}_{i,t}$ matrices is smaller
than the number of real entries in the matrix because of the
geometrical structure between the columns of the
matrix~\cite{4114278}.  Therefore $\tilde{\bV}_{i,t}$ could be paramaterized
using fewer parameters.
The orthonormal matrix $\tilde{\bV}_{i,t}$ with
$N_T$ rows and $N_R$ columns can be decomposed as
follows~\cite{4114278} (we suppress the $(i, t)$ subscripts in the
decomposition matrices)
\begin{equation}
\tilde{\bV}_{i,t} = \left[\prod_{k=1}^{N_{R}} \bD_{k} \left( \phi_{k,k},\ldots , \phi_{k,N_{R}} \right) \:  \prod_{l=1}^{N_{T}-k} \bG_{N_{T}-l,N_{T} -l+1} \big( \theta_{k,l}\big)  \right] \: \tilde{\bI}
\end{equation}
where $\bD_{k}$ is a diagonal matrix defined as
$\bD_{k}\big(\phi_{k,k}, \ldots, \phi_{k,N_T } \big)$ =
$\mbox{diag}\big( {\bf 1}_{k-1}, e^{j\phi_{k,k}},\ldots,
e^{j\phi_{k,N_T }} \big)$, ${\bf 1}_{k-1}$ represents $k-1$ ones,
and the $N_R \times N_T$ matrix $\tilde{\bI}$ =
$\big[\bI_{N_R }, {\bf 0}_{N_T ,N_T -N_R}\big]^{T}$. Here,
${\bf 0}_{N_T ,N_T -N_R }$ is an $N_T\times (N_T - N_R)$ matrix all of
whose elements are zero. Finally, we also have
\begin{equation}
\bG_{m-1,m}\big(\theta\big)  =
\begin{bmatrix}
\bI_{m-2} & & & \\
& \cos\theta & -\sin\theta & \\
& \sin\theta & \cos\theta & \\
& & & \bI_{t-m}
\end{bmatrix}
\end{equation}
Thus, for a $4 \times 2$ orthogonal matrix $\bV_{i,t}$, we have
\begin{align*}
  \bV_{i,t} & =
  \bD_{1}(\phi_{1,1},\ldots,\phi_{1,4})\bG_{3,4}(\theta_{1,1})
  \bG_{2,3}(\theta_{1,2}) \bG_{1,2}(\theta_{1,3})\times\\
& \bD_{2}(\phi_{2,2},\phi_{2,3},\phi_{2,4}) \bG_{3,4}(\theta_{2,1}) \bG_{2,3}(\theta_{2,2})\tilde{\bI}
\end{align*}
Here, $\phi_{k}$s are referred to as phases, and their collection for
the $i$-th subcarrier at time instant $t$ is captured in a vector
denoted by $\bf{\Phi_{i,t}}$. $\theta_{l}$s are called rotation angles, and
their collection is captured in a vector referred to as
$\bf{\Theta_{i,t}}$. We note that $\bf{\Phi_{i,t}}$ and $\bf{\Theta_{i,t}}$ represent
the phase and rotation angles for the $i$-th subcarrier at time
instant $t$, as represented in Fig.~\ref{fig:adpm-fig}. It is known
that $\phi_k \in (-\pi, \pi]$ is uniformly distributed between the two
extremes~\cite{4114278}, while $\theta_l$ is distributed
as~\cite{4114278}
\begin{equation}
2l(\sin\theta_l)^{2l-1}\cos\theta_l, \mbox{  }\theta_l \in \left[0, \frac{\pi}{2}\right) \mbox{, } l = 1,2,\ldots,N_T -1
\end{equation}
The given distribution of $\theta_l$ is used for non-uniform quantization
of $\bf{\Theta}$ during their initialization to reduce the quantization error.
In addition, $\phi_k$ and $\theta_l$ are statistically independent,
making this representation useful for tracking them separately across
time, as discussed below. The total number of
parameters obtained from the decomposition of a complex orthogonal
matrix $N_{T} \times N_{R} $ is $N_{R}(2N_{T} - N_{R})$. The number of
$\phi_k$ parameters is $N_{R}(2N_{T} - N_{R}-1)/2$ while the number of
$\theta$ parameters is $N_{R}(2N_{T} - N_{R}+1)/2$. Thus, $\bf{\Phi_{i,t}}$
is a vector of length $N_{R}(2N_{T} - N_{R}-1)/2$, while
$\bf{\Theta_{i,t}}$ has length $N_{R}(2N_{T} - N_{R}+1)/2$.~\cite{4114278}
\begin{figure}
\begin{center}
\includegraphics[width=0.8\columnwidth]{images/new-adpm.pdf}
\caption{\label{fig:adpm-fig}Subcarriers that need to be
  quantized. We see that the subcarrier indicated by
  red arrows has a precoder whose parameters are determined jointly
  using the neighbouring parameters, both in time and frequency.}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.72\columnwidth]{images/1024final_withtime}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ 1024
  subcarrier MIMO-OFDM system, with the ITU Cost207-RA channel profile
  for $f_dT_s = 3.41\times 10^{-6}$, corresponding to a velocity of 37.8 km/h.}
\label{fig:ber_ped}
\end{center}
\end{figure}

@@

Fig.~\ref{fig:ber_veh} shows a similar comparison for a situation with
a higher speed. The faster variation of the channel taps with time
necessitates the use of a smaller number of subcarriers (viz. 64
subcarriers), thereby permitting more accurate tracking of precoders
on a per subcarrier basis. As in the previous case, we observe that
the scalar parameter-based approach outperforms manifold based
quantization by about 1 dB for higher SNRs. Since, on a per subcarrier
basis, the channel variation is slower, the precoder tracking is
better than the previous case. Naturally, the rate of channel
variation is quite fast, which limits the efficacy of the proposed
method in tracking the channel parameters.
\begin{figure}
\begin{center}
\includegraphics[width=0.72\columnwidth]{images/64final_withtime}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ 64
  subcarrier MIMO-OFDM system, with the ITU Cost207-RA channel profile
  for $f_dT_s = 1.56\times 10^{-5}$, corresponding to a velocity of 172.8 km/h.}
\label{fig:ber_veh}
\end{center}
\end{figure}

>>>>

An underlying challenge in framing the predictive quantization and interpolation algorithms has been the high dimensional unitary structure of the SVD precoders. A large body of works utilizes the manifold geometry endowed by the unitary structure to frame such algorithms directly on the non-linear manifold ~\cite{mondal2007quantization,schwarz2013adaptive,5946308,6891198,Gupt1905:Predictive,pitaval2013codebooks}. On the other hand, there are numerous works which utilize scalar parametrization of the unitary precoders, given by Givens rotations and Householder transformations \cite{4114278,4556174,lou2013comparison}. While the manifold based methods offer a clean solution and have been shown to be very effective in reducing the feedback overhead, they are analytically terse and
involve operations over higher dimensional manifolds. The scalar parametrization algorithms enable usage of the traditional algorithms to predict, interpolate and quantize over vector spaces, and hence are analytically simpler.
Such methods have also been implemented in 802.11 ac and ax standards
\cite{lou2013comparison,ieee80211}. We are not aware of performance comparisons
between these two bodies of works, thereby making it difficult to
quantify from past work whether performance gains/losses can accrue by operating over scalar parametrization instead of manifolds. In this paper, we propose a joint time-frequency interpolation approach that builds upon the existing parametric quantization techniques. One of the key results of our work is that the proposed scalar parametrization method actually outperforms an equivalent manifold based technique \cite{Gupt1905:Predictive}, at the same time being analytically more tractable.

@@

\section{Scalar joint time-frequency scheme}
\subsection{Scalar Parametrization}
\label{givens}
The degrees of freedom in the $\tilde{\bV}_{i,t}$ matrices is smaller
than the number of real entries in the matrix because of the
geometrical structure between the columns of the
matrix~\cite{4114278}.  Therefore $\tilde{\bV}_{i,t}$ could be paramaterized
using fewer parameters.
The orthonormal matrix $\tilde{\bV}_{i,t}$ with
$N_T$ rows and $N_R$ columns can be decomposed as
follows~\cite{4114278} (we suppress the $(i, t)$ subscripts in the
decomposition matrices)
\begin{equation}
\tilde{\bV}_{i,t} = \left[\prod_{k=1}^{N_{R}} \bD_{k} \left( \phi_{k,k},\ldots , \phi_{k,N_{R}} \right) \:  \prod_{l=1}^{N_{T}-k} \bG_{N_{T}-l,N_{T} -l+1} \big( \theta_{k,l}\big)  \right] \: \tilde{\bI}
\end{equation}
where $\bD_{k}$ is a diagonal matrix defined as
$\bD_{k}\big(\phi_{k,k}, \ldots, \phi_{k,N_T } \big)$ =
$\mbox{diag}\big( {\bf 1}_{k-1}, e^{j\phi_{k,k}},\ldots,
e^{j\phi_{k,N_T }} \big)$, ${\bf 1}_{k-1}$ represents $k-1$ ones,
and the $N_R \times N_T$ matrix $\tilde{\bI}$ =
$\big[\bI_{N_R }, {\bf 0}_{N_T ,N_T -N_R}\big]^{T}$. Here,
${\bf 0}_{N_T ,N_T -N_R }$ is an $N_T\times (N_T - N_R)$ matrix all of
whose elements are zero. Finally, we also have
\begin{equation}
\bG_{m-1,m}\big(\theta\big)  =
\begin{bmatrix}
\bI_{m-2} & & & \\
& \cos\theta & -\sin\theta & \\
& \sin\theta & \cos\theta & \\
& & & \bI_{t-m}
\end{bmatrix}
\end{equation}
Thus, for a $4 \times 2$ orthogonal matrix $\bV_{i,t}$, we have
\begin{align*}
  \bV_{i,t} & =
  \bD_{1}(\phi_{1,1},\ldots,\phi_{1,4})\bG_{3,4}(\theta_{1,1})
  \bG_{2,3}(\theta_{1,2}) \bG_{1,2}(\theta_{1,3})\times\\
& \bD_{2}(\phi_{2,2},\phi_{2,3},\phi_{2,4}) \bG_{3,4}(\theta_{2,1}) \bG_{2,3}(\theta_{2,2})\tilde{\bI}
\end{align*}
Here, $\phi_{k}$s are referred to as phases, and their collection for
the $i$-th subcarrier at time instant $t$ is captured in a vector
denoted by $\bf{\Phi_{i,t}}$. $\theta_{l}$s are called rotation angles, and
their collection is captured in a vector referred to as
$\bf{\Theta_{i,t}}$. We note that $\bf{\Phi_{i,t}}$ and $\bf{\Theta_{i,t}}$ represent
the phase and rotation angles for the $i$-th subcarrier at time
instant $t$, as represented in Fig.~\ref{fig:adpm-fig}. It is known
that $\phi_k \in (-\pi, \pi]$ is uniformly distributed between the two
extremes~\cite{4114278}, while $\theta_l$ is distributed
as~\cite{4114278}
\begin{equation}
2l(\sin\theta_l)^{2l-1}\cos\theta_l, \mbox{  }\theta_l \in \left[0, \frac{\pi}{2}\right) \mbox{, } l = 1,2,\ldots,N_T -1
\end{equation}
The given distribution of $\theta_l$ is used for non-uniform quantization
of $\bf{\Theta}$ during their initialization to reduce the quantization error.
In addition, $\phi_k$ and $\theta_l$ are statistically independent,
making this representation useful for tracking them separately across
time, as discussed below. The total number of
parameters obtained from the decomposition of a complex orthogonal
matrix $N_{T} \times N_{R} $ is $N_{R}(2N_{T} - N_{R})$. The number of
$\phi_k$ parameters is $N_{R}(2N_{T} - N_{R}-1)/2$ while the number of
$\theta$ parameters is $N_{R}(2N_{T} - N_{R}+1)/2$. Thus, $\bf{\Phi_{i,t}}$
is a vector of length $N_{R}(2N_{T} - N_{R}-1)/2$, while
$\bf{\Theta_{i,t}}$ has length $N_{R}(2N_{T} - N_{R}+1)/2$.~\cite{4114278}
\begin{figure}
\begin{center}
\includegraphics[width=0.7\columnwidth]{images/new-adpm.pdf}
\caption{\label{fig:adpm-fig}Subcarriers that need to be
  quantized. We see that the subcarrier indicated by
  red arrows has a precoder whose parameters are determined jointly
  using the neighbouring parameters, both in time and frequency.}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.71\columnwidth]{images/1024final_withtime}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ 1024
  subcarrier MIMO-OFDM system, with the ITU Cost207-RA channel profile
  for $f_dT_s = 3.41\times 10^{-6}$, corresponding to a velocity of 37.8 km/h.}
\label{fig:ber_ped}
\end{center}
\end{figure}

@@

Fig.~\ref{fig:ber_veh} shows a similar comparison for a situation with
a higher speed. The faster variation of the channel taps with time
necessitates the use of a smaller number of subcarriers (viz. 64
subcarriers), thereby permitting more accurate tracking of precoders
on a per subcarrier basis. As in the previous case, we observe that
the scalar parameter-based approach outperforms manifold based
quantization by about 1 dB for higher SNRs. Since, on a per subcarrier
basis, the channel variation is slower, the precoder tracking is
better than the previous case. Faster channel variations would impose
limits on the efficacy of the proposed method in tracking the channel
parameters.
\begin{figure}
\begin{center}
\includegraphics[width=0.71\columnwidth]{images/64final_withtime}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ 64
  subcarrier MIMO-OFDM system, with the ITU Cost207-RA channel profile
  for $f_dT_s = 1.56\times 10^{-5}$, corresponding to a velocity of 172.8 km/h.}
\label{fig:ber_veh}
\end{center}
\end{figure}
